### 2024-01-25 parts/counter.tex
writing to publish

%! TeX program = xelatex
%! lang = en-US
\section{Polynomials}

@@

\begin{lemma}
    \label{n-poly-n-poly:example}
    Let $P \in \Nat[\vec{X}]$. Then, $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{lemma}
\begin{proof}
    We prove it by remarking that $\NPoly$
    is stable under non-negative linear combinations
    and contains the functions ...
\end{proof}

\begin{example}
    \label{negative-but-npoly:ex}
    The polynomial $P(X) \defined X^2 -2X + 2$ is
    a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
    that maps $aw \mapsto |w|^2$, and $\varepsilon \to 0$.
    It is clear that $f \in \NPoly$, and that 
    $P$ is \kl{represented} by
    $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
\end{proof}

@@

Let us now introduce the counter example that was found trying to generalize
the original proof to non-commutative \kl{$\Rel$-rational series} of
\kl{polynomial growth}. The counter example will use three indeterminate, and
we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
polynomial with at most two indeterminate. In particular, the examples
appearing in \cite{KARH77} are not invalidated, as they all use at most two
indeterminates. Furthermore, the converse implication, that is, that
\kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is correct, and can
even be strengthened as we will see in \cref{corrected-version:thm}.

@@

\begin{lemma}
    \label{thm:counter-example}
    The polynomial $\BadPoly$ belongs to $\CoveredPoly$,
    but is not a \kl{$\Nat$-rational polynomial}.
\end{lemma}
\begin{proof}
    It is clear that $\BadPoly$ is \kl{non-negative} as a sum of two 
    squares. We can develop
    the expression of $\BadPoly$ to 
    obtain
    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
    The \kl{maximal monomials} of $P$
    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which have
    \kl{positive coefficients}.

@@

    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
    Because $L \cap \set{a,b}^*$ is notoriously not a regular language
    (\cref{regular:fact}), we
    conclude that $L$ is not a regular language.
    However, $L = f^{-1}(\set{0})$, which is a regular language
    because $f$ is a \kl{$\Nat$-rational series}
    (\cref{pre-image-regular:fact}).
\end{proof}

\begin{corollary}
    The result stated in \cite[Theorem 3.3]{KARH77}, restated
    in \cref{karh:thm}, is false
    for all polynomials with at least $3$ indeterminate.
\end{corollary}

>>>>

%! TeX program = xelatex
%! lang = en-US
\section{Polynomials}
\label{polynomials:sec}

@@

\begin{fact}
    \label{n-poly-n-poly:example}
    Let $P \in \Nat[\vec{X}]$. Then, $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{fact}

\begin{example}
    \label{negative-but-npoly:ex}
    Let $P \defined X^2 -2X + 2$. Then $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
    that maps $aw \mapsto |w|^2$, and $\varepsilon \to 0$.
    It is clear that $f \in \NPoly$, and that 
    $P$ is \kl{represented} by
    $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
\end{proof}

@@

Let us now introduce the counter example that was found trying to generalize
the original proof $\ZPoly$.
The counter example will use three indeterminates, and
we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
polynomial with at most two indeterminate. In particular, the examples
appearing in \cite{KARH77} are not invalidated, as they all use at most two
indeterminates. Furthermore, the converse implication, that is, that
\kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is correct, and can
even be strengthened as we will see in \cref{corrected-version:thm}.

@@

\begin{lemma}
    \label{thm:counter-example}
    The polynomial $\BadPoly$ belongs to $\CoveredPoly$,
    but is not a \kl{$\Nat$-rational polynomial}.
\end{lemma}
\begin{proof}
    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
    the expression of $\BadPoly$ to 
    obtain
    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
    The \kl{maximal monomials} of $P$
    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which have
    \kl{positive coefficients}.

@@

    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
    Because $L \cap \set{a,b}^*$ not a regular language,
    we
    conclude that $L$ is not a regular language.
    However, $L = f^{-1}(\set{0})$ is a regular language
    (\cref{pre-image-regular:fact}). 
\end{proof}

\begin{corollary}
    The result stated in \cite[Theorem 3.3]{KARH77}, restated
    in \cref{karh:thm}, is false
    when allowing at least $3$ indeterminates.
\end{corollary}

### 2024-01-25 parts/intro.tex
writing to publish

Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{kleene1956representation}, Monadic Second Order Logic ($\MSO$)
\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
Among regular languages, there exists a robust subclass sharing similar
characterizations, namely the \emph{star-free} regular languages. These are
obtained via counter-free automata \cite{mcnaughton1971counter}, represented by
star-free regular expressions \cite{schutzenberger1965finite}, definable in
First Order Logic ($\FO$) \cite{perrin1986first}, and recognized by aperiodic
monoids \cite{schutzenberger1965finite}. It is decidable whether a regular
language is aperiodic, thanks to the characterizations in terms of minimal
automaton and/or syntactic monoid.

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}) \cite[see,
e.g.][]{bojanczyk2019string}. In this paper, we will focus on \kl{polyregular
functions}, whose class is denoted $\Poly$, and that shares with regular
languages the characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers. As for
regular languages, the class $\SF$ of \intro{star-free polyregular function}
can equivalently be introduced in terms of first-order definability,
counter-free pebble transducers, and suitable restrictions of for-tranducers.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton.
Indeed, for sequential functions that admit a canonical object, aperiodicity is
decidable \cite{choffrut03}, and more recently, aperiodicity has been proven
decidable for rational functions
\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
regular functions, although it has been proven decidable when changing the
semantics of function equality to account for the \emph{origins} in the output
\cite{bojanczyk14}.

Another way to approach the problem of deciding star-free for polyregular
functions is to not restrict the computational power, nor change the semantics,
but consider restrictions of the output alphabet. In particular, there have
been successful results proven on \emph{unary polyregular functions}, also
known as \kl{$\Nat$-polyregular functions}
\cite{doueneau2021pebble,doueneau2022hiding}. Let us note that results on
\emph{unary output functions} often generalize to the slightly more general
class of functions with \emph{commutative output}, and that algorithms leverage
the absence of combinatorial information in the output. Furthermore,
\kl{$\Nat$-polyregular functions} benefit from theorems proven in the field of
rational series \cite{berstel2011noncommutative}, where they coincide with
weighted automata of polynomial ambiguity, rational series of polynomial
growth, and a notion of quantitative $\MSO$
\cite{berstel2011noncommutative,kreutzer2013,schutzenberger1962}. Although
promising, the restriction to unary output did not yield effective procedure to
decide the aperiodicity of polyregular functions.

Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} called \kl{$\Rel$-polyregular functions} and denoted
by $\ZPoly$ \cite{LOPEZ23b}.  These functions generalise \kl{$\Nat$-polyregular
functions} by allowing negative outputs, and therefore enabling a limited form
of backtracking. The decidability of aperiodicity for \kl{$\Rel$-polyregular
functions} relies on the construction of a canonical object called the
\intro{residual transducer}, the latter being essentially based on differences
between functions, crucially leveraging negative outputs.
If the proof method does not carry from $\Rel$-output to $\Nat$-output
functions, it was conjectured that the semantic characterization of \kl{star
free $\Rel$-polyregular functions} ($\ZSF$) would continue to characterize
\kl{star-free $\Nat$-polyregular functions} ($\NSF$), and in particular that
$\NPoly \cap \ZSF = \NSF$.

Let us remark that, \kl{$\Nat$-polyregular functions} and
\kl{$\Rel$-polyregular functions} are proper subsets respectively of
\kl{$\Nat$-rational series} ($\ZRat$) and of \kl{$\Rel$-rational series}
($\NRat$). For these larger classes of functions, several competing notions of
aperiodicity exist \cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b},
and decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ among $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved in the case
of \emph{polynomials} by \textcite{KARH77}.

@@

\begin{figure}
    \includestandalone[width=\linewidth]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rat$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
    }
    \label{previously-known-inclusions:fig}
\end{figure}

\paragraph{Contributions.} We prove that for \kl{commutative} functions, all
the membership problems involving the classes $\ZSF$, $\ZPoly$, $\NSF$, and
$\NPoly$ are all decidable (this is the lower left square of inclusions in
\cref{previously-known-inclusions:fig}), and effective conversion algorithms
are provided. Furthermore, we show that in the commutative case, the conjecture
stating that $\ZSF \cap \NPoly = \NSF$ holds.

These results are obtained by completely understanding which \emph{polynomials}
are computable using \kl{$\Nat$-polyregular functions}, and lifting these
results from polynomials to \kl{$\Rel$-polyregular functions}. This study
invalidates the main theorem of \cite{KARH77}, for which we provide an explicit
counter-example.

\begin{itemize}
    \item We show that commutativity is decidable.
    \item We provide a candidate semantic characterization
    \item We prove that it is correct for commutative inputs
    \item We provide a counter-example to the theorem of nanania.
    \item We show that effective procedures are possible for all arrows 
        and answer positively to the conjecture of XX
        in the case of commutative inputs.
\end{itemize}
It was conjectured in a thesis that some particular form of polynomials
would characterize star-free functions.

This paper follows the line of
research on \kl{polyregular functions} that focuses on semantic
characterizations of the asymptotic behavior, rather than the construction
of canonical objects of computation
\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}. The goal was to
decide arrows X,Y and Z, namely, to understand precisely the relationship
between star-free and $\Nat$ output. In the case of \kl{$\Nat$-rational
series}, a result of \citeauthor{KARH77} characterizes \kl{commutative}
\kl{$\Nat$-rational series} among \kl{$\Rel$-rational series}
\cite{KARH77}.
This conjecture was based on the previously introduced result of
\textcite[Theorem 3.3]{KARH77}, which we will disprove in \cref{sec:c-example}.
We then provide a corrected version of the theorem, and prove the conjecture of
Douéneau in the case of \emph{commutative} $\Nat$-rational series of polynomial
growth.

\cref{thm:counter-example}
\cref{corrected-version:thm}

\cref{decidable-n-poly:thm}
\cref{zsf-npoly-nsf:thm}

\hrule

\paragraph*{Outline of the paper.}
In \cref{preliminaries:sec}, we introduce the computational models
and prove that 
\begin{itemize}
    \item In a first section, we recall the preliminary definitions.
    \item In a second section, we provide a counter example to the
        theorem of XX.
    \item In a third section, we provide an alternative result.
    \item In a fourth section, we make the results effective,
        and compute everyting.
\end{itemize}

>>>>

Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{kleene1956representation}, Monadic Second Order Logic ($\MSO$)
\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
Among regular languages, there exists a robust subclass sharing similar
characterizations, namely the \emph{star-free} regular languages. These are
obtained via counter-free (minimal) automata \cite{mcnaughton1971counter},
represented by star-free regular expressions \cite{schutzenberger1965finite},
definable in First Order Logic ($\FO$) \cite{perrin1986first}, and recognized
by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
regular language is aperiodic, thanks to the characterizations in terms of
minimal automaton and/or syntactic monoid.

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}). In this paper,
we will focus on \kl{polyregular functions} ($\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite[see, e.g.][]{bojanczyk2019string}. As for regular languages, the class
$\SF$ of \intro{star-free polyregular function} can equivalently be introduced
in terms of first-order definability, counter-free pebble transducers, and
suitable restrictions of for-tranducers.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton.
Indeed, for sequential functions that admit a canonical object, aperiodicity is
decidable \cite{choffrut03}, and more recently, aperiodicity has been proven
decidable for rational functions
\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
regular functions, although it has been proven decidable when changing the
semantics of function equality to account for the \emph{origins} in the output
\cite{bojanczyk14}. To construct subclasses of \kl{polyregular function}, an
interesting parameter is the \kl{growth rate} of a function $f$, which is given
by a number $k \in \Nat$, which can be used to create a strict hierarchy inside
$\Poly$. We write $\Poly[k]$ for the \kl{polyregular functions} of \kl{growth
rate} $k$. It has been shown that the parameter $k$ controls the maximal
complexity of the $\MSO$ interpretation needed to represent the function
\cite{bojanczyk2022transducers}.

Another way to approach the problem of deciding if a \kl{polyregular function}
is \kl{star-free}, is to not restrict the computational power (via the
\kl{growth rate}), nor change the semantics (by adding origins), but consider
restrictions of the output alphabet. In particular, there have been successful
results proven on \emph{unary polyregular functions}, also known as
\kl{$\Nat$-polyregular functions} ($\NPoly$)
\cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
needed to compute $f$ \cite{doueneau2021pebble}, which does not hold on $\Poly$
\cite{bojanczyk2022transducers}. Let us remark that the restriction to
\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
outputs in the free commuative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{unary output} is really about
enforcing commutativity on the output.

Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\ZPoly$)
\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). If the proof method does not carry
from $\Rel$-output to $\Nat$-output functions, it was conjectured that the
semantic characterization of \kl{star-free $\Rel$-polyregular functions}
($\ZSF$) would continue to characterize \kl{star-free $\Nat$-polyregular
functions} ($\NSF$).

\begin{conjecture}[{\cite[Conjecture 7.61]{gaetanphd}}]
    \label{zsf-nsf:conjecture}
    Let $k \in \Nat$.
    A function $f \in \NPoly[k]$
    belongs to $\NSF[k]$ if and only if
    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
    In particular,
    $\NSF[k] = \ZSF[k] \cap \NPoly$.
\end{conjecture}

It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \kl{$\Nat$-rational series} ($\NRat$) (resp. of \kl{$\Rel$-rational series}
$\ZRat$), that have been characterized in terms of \kl{growth rate}, ambiguity,
and a notion of \emph{quantitative} $\MSO$
\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
theorems obtained on these \cite{berstel2011noncommutative}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}. The decidability of $\NPoly$ inside $\ZPoly$ (resp. $\NSF$
inside $\ZSF$) remains an open problem
\cite[see e.g.][Open question 5.55]{gaetanphd}.

@@

\begin{figure}
    \includestandalone[width=\linewidth]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
    }
    \label{previously-known-inclusions:fig}
\end{figure}

\paragraph{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}. Remark that this enforces a
commutativity property both on the input alphabet and on the output alphabet.
In this restricted setting, we answer positively to \cref{zsf-nsf:conjecture},
and prove that the membership problems between $\NSF$, $\ZSF$, $\NPoly$, and
$\ZPoly$ are all decidable, with effective conversion algorithms. This
effectively shows that all arrows in the lower left square of
\cref{previously-known-inclusions:fig} are decidable in the \kl{commutative}
setting. Furthermore, we characterize \kl{commutative} \kl{$\Nat$-polyregular
function} as the class of \kl{$\Rel$-polyregular functions} that are
\kl{combinatorial} (defined page \kpageref{combinatorial}), which is a semantic
property akin to the one \kl{ultimate polynomiality} that characterizes $\ZSF$.
As a byproduct of our analysis, we provide a counter example to the
characterization of polynomials computed by \kl{$\Nat$-rational series} given
by \textcite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.

Finally, we make a first step towards understanding the non-commutative by
introducing a canonical model of computation for \kl{$\Nat$-polyregular
functions} built upon residuals, called the \kl{residual transducer}. This
allows the characterization of $\NPoly[k]$ as the class of functions $f$ such
that the collection $\Res(f)$ of all \kl{residuals} of $f$ is a
\kl{well-quasi-ordering} with respect to the partial ordering $g \npolyleq[k]
h$ if and only if $h - g \in \NPoly[k-1]$. In this sense, it generalizes to
orderings the characterization of $\ZPoly[k]$ as the class of functions $f$
such that $\Res(f)$ has finite index for the equivalence relation $g
\zpolyequiv[k] h$ if and only if $g - h \in \ZPoly[k-1]$. The \kl{residual
transducer} is effectively computable if $f$ is a \kl{commutative}
\kl{$\Nat$-polyregular function}, and it is conjectured to be computable for
all functions in $\NPoly$. Finally, we prove that for $f \in \NPoly[0]$, $f \in
\NSF$ if and only if the \kl{residual transducer} of $f$ has no \kl{counters},
and conjecture that in full generality, $f \in \NSF$ if and only if the
\kl{residual transducer} of $f$ is \kl{counter-free}.

\paragraph{Outline of the paper.} In \cref{preliminaries:sec}, we introduce the
multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decide-commutative:lemma}), and give the
basic definitions used on multivariate polynomials. In \cref{polynomials:sec},
we focus on \emph{polynomials}, by introducing the \cref{karh:thm} of
\textcite{KARH77}, providing a counter-example in \cref{thm:counter-example},
and an alternative characterization in \cref{corrected-version:thm}. In
\cref{beyond-polynomials:sec}, we leverage the results over polynomials to
decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture} holds
for \kl{commutative} functions (\cref{zsf-npoly-nsf:thm}). Finally, in
\cref{beyond-commutative:sec}, we introduce the notion of \kl{residual
transducer} (\cref{residual-transducer:def}), and prove in
\cref{non-commutative-npoly:thm} that it characterizes $\NPoly$ inside
$\ZPoly$.
